---
title: "OPI Forecasts 2026"
author: Thomas Buehrens, WDFW (tbuehrens@dfw.wa.gov), Mark Sorel, WDFW (mark.sorel@dfw.wa.gov), Cassandra Leeman, ODFW (cassandra.r.leeman@odfw.oregon.gov), and Shannon Conley, WDFW (shannon.conley@dfw.wa.gov)
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"https://privatelands.wdfw.wa.gov/wdfwlogo_clrnotxt.png"\" style=\"float: right;width: 150px;\"/>')
   });
</script>
```
```{=html}
<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"https://raw.githubusercontent.com/tbuehrens/Salmon_Inseason_Forecasts/test/data/ODFW_logo.png"\" style=\"float: right;width: 150px;\"/>')
   });
</script>
```

------------------------------------------------------------------------

Last Updated `r format(Sys.time(), '%m/%d/%Y')`.

------------------------------------------------------------------------

# Overview

This report details the process of fitting Auto-Regressive Integrated
Moving Average (ARIMA) models to forecast OPI Coho salmon runsize. The
model functions are flexibly configured to allow the user to supply
total runsize data and a series of covariates.

```{r set_options, echo = TRUE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE)
set.seed(123)
```

<!-- We also need a couple of helper functions which we will define -->

```{r load_funcs, message = FALSE, warning = FALSE,results = "hide"}
wd_functions<-"functions"
sapply(FUN = source, paste(wd_functions, list.files(wd_functions), sep="/"))
```

<!-- Here we will load & install packages we need to use (needs internet connection if packages not already installed) -->

```{r load_packages, message = FALSE, warning = FALSE,results = "hide"}
packages_list<-c("tidyverse"
                 ,"forecast"
                 ,"mgcv"
                 ,"ggplot2"
                 ,"MASS"
                 ,"RColorBrewer"
                 ,"kableExtra"
                 ,"modelr"
                 ,"kableExtra"
                 ,"reshape2"
                 ,"ggfortify"
                 ,"clock"
                 ,"smooth"
                 ,"scales"
                 ,"gtools"
                 # ,"CombMSC"
                 ,"here"
                 ,"MuMIn"
                 ,"janitor"
                 ,"rvest"
                 ,"lubridate"
                 ,"rnoaa"
                 ,"ncdf4"
                 ,"magrittr"
                 ,"ggcorrplot"
                 ,"parallel"
                 ,"foreach"
                 ,"doParallel"
                 ,"readxl"
                 )
install_or_load_pack(packages = packages_list)

 if(!require(SalmonForecasting)|(packageVersion("SalmonForecasting")!="0.1.0")){
   devtools::install_github("wdfw-fp/salmonforecast@v0.1.0")
   }


# pacman::p_load(tidyverse
#                  ,forecast
#                  ,mgcv
#                  ,ggplot2
#                  ,MASS
#                  ,RColorBrewer
#                  ,kableExtra
#                  ,modelr
#                  ,kableExtra
#                  ,reshape2
#                  ,ggfortify
#                  ,clock
#                  ,smooth
#                  ,scales
#                  ,gtools
#                  ,CombMSC
#                  ,here
#                  ,MuMIn
#                  ,janitor
#                  ,rvest
#                  ,lubridate
#                  ,rnoaa
#                  ,ncdf4
#                  ,magrittr
#                  ,ggcorrplot
#                  ,parallel
#                  ,foreach
#                  ,doParallel
#                  ,readxl)

```

## User Inputs

This section contains parameters input to control the forecasting
process

```{r user_inputs_data, message = FALSE, warning = FALSE,results = "show"}
#=========
# Raw Data
#=========
yr_start<-1967
yr_end<-year(Sys.Date())
#dam="BON"
species="Coho"
#===========================
# Summarization for analysis
#===========================
date_start_analysis<-ymd("1967/1/1") #this will be the first date of data used in the analysis (and this month/day first in years thereafter)
date_end_analysis<-ymd("2026/12/31") #this will be the last date of data used in the analysis (and this month/day last in years preceding)
forecast_period_start_m<-1 #this will be the month associated with the first month/day of the seasonal estimate period each year...can be after first data used to estimate it or the same
forecast_period_start_d<-1 #this will be the month day associated with the first month/day of the seasonal estimate period each year...can be after first data used to estimate it or the same
last_data<-Sys.Date()
#==================
#forecasting params
#==================
leave_yrs<- 31
TY_ensemble<-16
k<-1
                                                                                            covariates<-c(#new
                                                                                                          "lag1_log_JackOPI"
                                                                                                          ,"lag1_log_SmAdj"
                                                                                                          ,"lag1_NPGO"
                                                                                                          ,"lag1_PDO"
                                                                                                          ,"WSST_A"
                                                                                                          
                                                                                                          #OCN
                                                                                                          ,"PDO.MJJ"
                                                                                                          ,"MEI.OND"
                                                                                                          ,"UWI.JAS"
                                                                                                          ,"SST.AMJ"
                                                                                                          ,"SSH.AMJ"
                                                                                                          ,"UWI.SON"
                                                                                                          
                                                                                                          #Very correlated with variables already included
                                                                                                          #,"lag1_fall_Nino3.4"
                                                                                                          #,"SST.J"
                                                                                                          
                                                                                                          #Original OPIT indicators very correlated with variables already used(but untransformed)
                                                                                                          #"lag1_JackOPI"
                                                                                                          #,"lag1_SmAdj"
                                                                                                          
                                                                                                          #NOAA Ocean Indicators
                                                                                                          #,"lag1_sp_phys_trans"
                                                                                                          #,"lag1_PC1"
                                                                                                          ) 
plot_results = F
first_forecast_period = 1
write_model_summaries = TRUE
find_best=T
redo_forecasts=FALSE #where to do one-year-ahead forecasts for all models and years if there is a file saved with some. Change to TRUE if changing the min or max number of variables allowed per model, the covariates, or adding a new year of data. TO-DO: add functionality to only make projections for years or models that don't exist and get rid of any that you don't want (e.g., if you reduce max number of covariates per model)
#==============
#Ensemble Params
#===============
min_vars<-0
max_vars<-6
forecast_type<-"preseason"
stack_metric<-"MAPE"
num_models<-10
rolling_year_window<-15
alpha<-0.3 # exponential smoothing of APE calculations

inputs<-list(
  yr_start = yr_start,
  yr_end = yr_end,
  species = species,
  date_start_analysis = date_start_analysis,
  date_end_analysis = date_end_analysis,
  forecast_period_start_m = forecast_period_start_m,
  forecast_period_start_d = forecast_period_start_d,
  last_data = last_data,
  leave_yrs = leave_yrs,
  TY_ensemble = TY_ensemble,
  k = k,
  covariates = covariates,
  plot_results = plot_results, 
  first_forecast_period = first_forecast_period,
  write_model_summaries = write_model_summaries, 
  find_best = find_best,
  min_vars = min_vars,
  max_vars = max_vars,
  forecast_type = forecast_type,
  stack_metric = stack_metric,
  num_models = num_models,
  rolling_year_window=rolling_year_window,
  redo_forecasts=redo_forecasts,
  alpha=alpha
)

print(inputs)
```

## Get Raw Data

Here we load the response variable (observed OPI coho abundance) and
covariates used in past forecasts from Erik Suring's GitHub repository
[**OPIH Evaluation**](https://github.com/ErikSuring/OPIH_Evaluation). We
also load covariates used in the Oregon Coast Natural (OCN) coho
forecast, as well as ancillary additional covariates. More details are
available in the appendix.

```{r get_data, message=FALSE, warning=FALSE, results="show"}

make_dat<-function(file_path){

if(file.exists(file_path)){
  return(read_csv(here::here(file_path)))
  
}else{
#OPIHData <- read.table("https://raw.githubusercontent.com/ErikSuring/OPIH_Evaluation/main/PUB2023.txt", header = TRUE, sep = "\t", strip.white = TRUE, comment.char = "#")
# OPIHData<-read.table(here::here("data/PUB2024_manual_update.txt"), header = TRUE, sep = "\t", strip.white = TRUE, comment.char = "#")
OPIHData<-readxl::read_excel(here::here("data/covars_jan2025_MS.xlsx")) |> as.data.frame()

if(max(OPIHData$Year)<year(Sys.Date())){
  maxyear=max(OPIHData$Year)+1
  OPIHData%<>%
    right_join(tibble(Year=c(unique(OPIHData$Year),year(Sys.Date()))))
}

nYears <- length(OPIHData[,1])
earliestYear <- OPIHData[1,1]
latestYear <- OPIHData[nYears,1]

OPIHData$AdltAll <- c(OPIHData$AdltSRS[1:17],OPIHData$AdltMSM[18:nYears])   # row 18  <-  1986. Use SRS data prior, MSM data post
OPIHData$AdltAllno83 <- OPIHData$AdltAll
OPIHData$AdltAllno83[15] <-  NA   #  1983 is removed from model input data
OPIHData[,"AdltAll"] <- OPIHData$AdltAll
OPIHData[,"AdltAllno83"] <- OPIHData$AdltAllno83

# Lag Jack and Smolt data to return year
OPIHData[,"lagJackCR"] <- c(NA, OPIHData$JackCR[1:nYears-1])
OPIHData[,"lagJackOC"] <- c(NA, OPIHData$JackOC[1:nYears-1])
OPIHData[,"lagJackOPI"] <- OPIHData$lagJackCR + OPIHData$lagJackOC

OPIHData[,"lagSmD"] <- c(NA, OPIHData$SmD[1:nYears-1])
OPIHData[,"lagSmCR"] <- c(NA, OPIHData$SmCR[1:nYears-1])
OPIHData[,"lagSmAdj"] <- OPIHData$lagJackCR*(OPIHData$lagSmD/OPIHData$lagSmCR)   #  Calculate Smolt Adjustment

# Output OPIHData as a .csv file
#write.table(OPIHData, file = "OPIHData.csv", sep = ", ", row.names=FALSE)

OPIHData%<>%
  dplyr::rename(year=Year,
                abundance = AdltAllno83
                )%>%
  as_tibble()%>%
  arrange(year)

Yrlist<-data.frame(year=c(min(OPIHData$year):(maxyear)))

OPIHData%<>%
  right_join(Yrlist)%>%
  arrange(year)

#=========================================================
#get PDO data
#=========================================================
PDO<-read_table("https://psl.noaa.gov/pdo/data/pdo.timeseries.ersstv5.csv",skip=1,col_names=F,comment="#")%>%
  dplyr::rename(Date=X1,PDO=X2)%>%
  filter(!PDO < -99)%>%
  mutate(Date=as.Date(Date),Month=month(Date),Year=as.integer(year(Date)))%>%
  group_by(Year)%>%
  add_tally()%>%
  #filter(!Month>6)%>% #use only spring (Jan-June) NPGO
  #filter(!n < 12)%>% #use only complete years
  group_by(Year)%>%
  dplyr::rename(year=Year)%>%
  dplyr::summarise(PDO=mean(PDO))%>%
  right_join(Yrlist)%>%
  mutate(lag1_PDO = lag(PDO,1))%>%
  dplyr::select(year,lag1_PDO)%>%
  filter(!is.na(lag1_PDO))
#=========================================================
#get NPGO data
#=========================================================



NPGO<-read_table("https://www.o3d.org/npgo/data/NPGO.txt",skip=29,col_names=F,comment="#")%>%
  filter(!is.na(X2))%>%
  dplyr::rename(Year=X1,Month=X2,NPGO=X3)%>%
  mutate(Year=as.integer(as.numeric(Year)))%>%
  group_by(Year)%>%
  add_tally()%>%
  #filter(!Month>6)%>% #use only spring (Jan-June) NPGO
  #filter(!n < 12)%>% #use only complete years
  group_by(Year)%>%
  dplyr::summarise(NPGO=mean(NPGO),n=n())%>%
  dplyr::rename(year=Year)%>% 
  right_join(Yrlist)%>%
  mutate(lag1_NPGO = lag(NPGO),lag_n=lag(n))%>%

  arrange(year)%>%
  dplyr::select(year,lag1_NPGO,lag_n)%>%
  filter(!is.na(lag1_NPGO))
#=========================================================
#get NOAA indicator data, wrangle into usable format, plot
#=========================================================
# if(!file.exists(here::here("data/2023stoplight.xlsx"))){
#   url<-"https://www.fisheries.noaa.gov/s3/2024-01/2023-Stoplight-RAWDATA-FINAL-USE-THIS-ONE.xlsx"
#   download.file(url, destfile = here::here("data/2023stoplight.xlsx"), method = "auto", mode = "wb")
#   indicators<-readxl::read_xlsx(here::here("data/2023stoplight.xlsx"))
# }else{
#   indicators<-readxl::read_xlsx(here::here("data/2023stoplight.xlsx"))
# }
# 
# indicators<-indicators%>%
#   filter(!is.na(`Ecosystem Indicators`))%>%
#   mutate(across(c(starts_with("1"),starts_with("2")),~as.numeric(.)))%>%
#   pivot_longer(names_to = "Year",
#                 cols=c(starts_with("1"),starts_with("2")),
#                 values_to = "value")%>%
#   pivot_wider(names_from=`Ecosystem Indicators`,values_from=value)%>%
#   mutate(year=as.integer(Year))%>%
#   dplyr::select(-Year)%>%
#   right_join(Yrlist)%>%
#   arrange(year)%>%
#   mutate(lag1_PC1 = lag(scale(`Principal Component scores (PC1)`)[,1]))%>%
#   filter(!is.na(lag1_PC1))
#=========================================
# Get ENSO index
#=========================================
# enso<-read_table("https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/nino34.long.anom.data",skip=1,col_names = F)%>%
#   as_tibble()%>%
#   setNames(c("year",1:12))%>%
#   filter(year%in%as.character(yr_start:yr_end))%>%
#   mutate(across(everything(),~as.numeric(.)))%>%
#   pivot_longer(names_to = "month",values_to = "Nino3.4",cols=c(!year))%>%
#   filter(Nino3.4>-10 & Nino3.4 <10)%>%
#   filter(month%in%c(10:12))%>%
#   group_by(year)%>%
#   summarise(fall_Nino3.4=mean(Nino3.4))%>%
#   right_join(Yrlist)%>%
#   mutate(lag1_fall_Nino3.4=lag(fall_Nino3.4))%>%
#   dplyr::select(-fall_Nino3.4)

#===========================================================================================================
#Get ERSST data: get_ersst_v5_data takes A LONG TIME (1 hr) vs. get_ersst_v5_data_V2 which is much quicker!
#==========================================================================================================
# sstdat<-get_ersst_v5_data_V2(years=c(min(Yrlist$year):max(Yrlist$year)),
#                              data.dir="https://stateofwa.sharepoint.com/:u:/r/sites/DFW-TeamWDFWOPIModelReview/Shared%20Documents/General/sst.mnmean.nc?csf=1&web=1&e=mhx1bc",
#                                #"C:\\Users\\sorelmhs\\Washington State Executive Branch Agencies\\DFW-Team WDFW OPI Model Review - General" ,
#                              ncfilename="sst.mnmean.nc",
#                              latrange=c(44,50),
#                              lonrange=c(-125,-120)
#                              )

sstdat<-get_ersst_v5_new(years=c(min(Yrlist$year):max(Yrlist$year)),
                           data.dir=here::here("data"),
                           ncfilename="sst.mnmean_1_27_2026.nc",
                           latrange=c(44,50),
                           lonrange=c(-125,-120)
)%>%
  filter(!is.na(sst))%>%
  group_by(year,month)%>%
  summarize(meanSST=mean(sst,na.rm=T),n=n())%>%
  mutate(year=as.numeric(year),
         month=as.numeric(month)
         )%>%
  bind_rows(read_csv(here::here("data/January_ERSST_v5_2026.csv"))%>%
              dplyr::select(year,month,meanSST=SST)

            )%>%
  group_by(month)%>%
  mutate(gmeanSST=mean(meanSST,na.rm=T))%>%
  ungroup()%>%
  mutate(resid=meanSST-gmeanSST)


ssta<-sstdat%>%
  dplyr::select(year,month,resid)%>%
  mutate(month=as.numeric(month))%>%
  mutate(year=ifelse(month>9,year+1,year))%>%
  group_by(year)%>%
  summarise(WSST_A = mean(resid[month>9|month<2]),
            SSST_A = mean(resid[month<=9|month>=4]),
            )

#================
#Plot ERSST data
#================
# ggplot(sstdat,aes(x=factor(month),y=meanSST,group=year))+
#   geom_hline(yintercept=0,linetype="dashed")+
#   scale_color_brewer(palette = "Spectral")+
#   geom_line()+
#   facet_wrap(~factor(year))
# 
# ggplot(sstdat,aes(x=factor(month),y=resid,group=year))+
#   geom_hline(yintercept=0,linetype="dashed")+
#   scale_color_brewer(palette = "Spectral")+
#   geom_line()+
#   facet_wrap(~factor(year))
# 
# ggplot(sstdat,aes(x=factor(month),y=sstdiff,group=year))+
#   geom_hline(yintercept=0,linetype="dashed")+
#   scale_color_brewer(palette = "Spectral")+
#   geom_line()+
#  facet_wrap(~factor(year))
#==========================================
#Get OCN covariates
#==========================================
OCN<-readxl::read_excel("data/OCNR_Forecast_Covar.xlsx")%>%
  # as_tibble()%>%
  dplyr::rename(year=YEAR)#%>%
  # dplyr::select(-c("ADULTS","SPAWNERS","Spawners"))
#================================================================
dat<-OPIHData%>%
  right_join(Yrlist)%>%
  left_join(PDO)%>%
  left_join(NPGO)%>%
  # left_join(indicators)%>%
  # left_join(enso)%>%
  left_join(ssta)%>%
  left_join(OCN)%>%
  dplyr::rename(lag1_JackOPI = lagJackOPI,
                lag1_SmAdj = lagSmAdj
                )%>%
  mutate(species = "Coho",
         period = 1,
         lag1_log_JackOPI = log(lag1_JackOPI),
         lag1_log_SmAdj = log(lagJackCR) * (lagSmD/lagSmCR),
         lag1_sp_phys_trans = lag( "Physical Spring Trans.\r\nUI based (day of year)")
         )%>%
  ungroup()%>%
  dplyr::select(year,species,period,abundance,all_of(unique(unlist(covariates))))%>%
  filter(
    across(
      .cols = all_of(unique(unlist(covariates))),
      .fns = ~ !is.na(.x)
    )
  )#%>%mutate(across(!c(year,species,period,abundance),boxcox_scale(.)))
 write.csv(dat,here::here("data/dat.csv"))
 return(read_csv(here::here(file_path)))
 }#closes Else
  
}#closes makedat function

dat<-make_dat("data/dat.csv")

dat%>%
  dplyr::select(-c(year,species,period,`...1`))%>%
  cor(use="pairwise.complete.obs")%>%
  ggcorrplot(hc.order = FALSE, type = "lower", outline.col = "white", p.mat = NULL, sig.level = 0.05,method="circle")



# ggsave(filename="corplot.png")

# 
# dat%>%
#   dplyr::select(-c(species,period))%>%
#   pivot_longer(abundance:UWI.SON) |>
#   group_by(name) |> 
#   mutate(value=scale(value),
#          this_year=(year==2026)) |> 
#     # filter(year>=2015) |> 
#   ggplot(aes(x=year,y=value,fill=this_year))+geom_col()+facet_wrap(~name)+ylab("Z score")+theme(axis.title.x = element_blank(),legend.position = "none")+scale_fill_manual(values=c("black","firebrick3"))+scale_x_continuous(breaks=c(1975,2000,2025))
# 
# ggsave(filename=here::here("covariates_all_yrs.png"))


# dat%>%
#   dplyr::select(-c(species,period))%>%
#   pivot_longer(abundance:UWI.SON) |>
#   group_by(name) |> 
#   mutate(value=scale(value),
#          this_year=(year==2026)) |> 
#     filter(year>=2015) |>
#   ggplot(aes(x=year,y=value,fill=this_year))+geom_col()+facet_wrap(~name)+ylab("Z score")+theme(axis.title.x = element_blank(),legend.position = "none")+scale_fill_manual(values=c("black","firebrick3"))+scale_x_continuous(breaks=c(2015,2020,2025))
# 
# ggsave(filename=here::here("covariates_recent.png"))
#   
# 
# dat%>%
#   dplyr::select(-c(species,period))%>%
#   pivot_longer(abundance:UWI.SON) |>
#
#   group_by(name) |>
#   mutate(value=scale(value),
#          this_year=(value>0)) |>
#     # filter(year>=2015) |>
#   ggplot(aes(x=year,y=value,fill=this_year))+geom_col()+facet_wrap(~name)+ylab("Z score")+theme(axis.title.x = element_blank(),legend.position = "none")+scale_fill_manual(values=c("darkblue","firebrick3"))+scale_x_continuous(breaks=c(2075,2000,2025))

# ggsave(filename=here::here("covariates2_all_yers.png"))




# dat%>%
#   dplyr::select(-c(species,period))%>%
#   pivot_longer(abundance:UWI.SON) |>
# 
#   group_by(name) |>
#   mutate(value=scale(value),
#          this_year=(value>0)) |>
#     filter(year>=2015) |>
#   ggplot(aes(x=year,y=value,fill=this_year))+geom_col()+facet_wrap(~name)+ylab("Z score")+theme(axis.title.x = element_blank(),legend.position = "none")+scale_fill_manual(values=c("darkblue","firebrick3"))+scale_x_continuous(breaks=c(2015,2020,2025))

# ggsave(filename=here::here("covariates2_recent.png"))

# 
#   cor(use="pairwise.complete.obs")%>%
#   ggcorrplot(hc.order = FALSE, type = "lower", outline.col = "white", p.mat = NULL, sig.level = 0.05,method="circle")
# 
# 


#read in OPIH data (again) to use for evaluating the current (as of 2022) forecasting methodology
{
# OPIHData<-read.table(here::here("data/PUB2024_manual_update.txt"), header = TRUE, sep = "\t", strip.white = TRUE, comment.char = "#")
OPIHData<-readxl::read_excel(here::here("data/covars_jan2025_MS.xlsx")) |> as.data.frame()


if(max(OPIHData$Year)<year(Sys.Date())){
  maxyear=max(OPIHData$Year)+1
  OPIHData%<>%
    right_join(tibble(Year=c(unique(OPIHData$Year),year(Sys.Date()))))
}

nYears <- length(OPIHData[,1])
earliestYear <- OPIHData[1,1]
latestYear <- OPIHData[nYears,1]

OPIHData$AdltAll <- c(OPIHData$AdltSRS[1:17],OPIHData$AdltMSM[18:nYears])   # row 18  <-  1986. Use SRS data prior, MSM data post
OPIHData$AdltAllno83 <- OPIHData$AdltAll
OPIHData$AdltAllno83[15] <-  NA   #  1983 is removed from model input data
OPIHData[,"AdltAll"] <- OPIHData$AdltAll
OPIHData[,"AdltAllno83"] <- OPIHData$AdltAllno83

# Lag Jack and Smolt data to return year
OPIHData[,"lagJackCR"] <- c(NA, OPIHData$JackCR[1:nYears-1])
OPIHData[,"lagJackOC"] <- c(NA, OPIHData$JackOC[1:nYears-1])
OPIHData[,"lagJackOPI"] <- OPIHData$lagJackCR + OPIHData$lagJackOC

OPIHData[,"lagSmD"] <- c(NA, OPIHData$SmD[1:nYears-1])
OPIHData[,"lagSmCR"] <- c(NA, OPIHData$SmCR[1:nYears-1])
OPIHData[,"lagSmAdj"] <- OPIHData$lagJackCR*(OPIHData$lagSmD/OPIHData$lagSmCR)   #  Calculate Smolt Adjustment

# Output OPIHData as a .csv file
#write.table(OPIHData, file = "OPIHData.csv", sep = ", ", row.names=FALSE)

OPIHData%<>%
  dplyr::rename(year=Year,
                abundance = AdltAllno83
                )%>%
  as_tibble()%>%
  arrange(year)

Yrlist<-data.frame(year=c(min(OPIHData$year):(maxyear)))

OPIHData%<>%
  right_join(Yrlist)%>%
  arrange(year)
  }
```



## Analysis and summarization

This section shows the code used to analyze the data and summarize the
results

```{r Analysis_v2, message=FALSE, warning=FALSE, results="hide"}

# test_pack<-SalmonForecasting::do_forecast(dat,max_vars = max_vars,n_cores =8,ts_freq=2,seasonal = TRUE)
# 
# test_pack$plots_and_tables$Table2

if(find_best ==T){
   best_covariates<-SalmonForecasting::all_subsets(series=dat,covariates=covariates,min=min_vars,max=max_vars,type=forecast_type,fit=FALSE)
  saveRDS(best_covariates,"best_covariates_OPI_preseason.rds")
}

best_covariates<-readRDS("best_covariates_OPI_preseason.rds")

# Table1 <- best_covariates[[2]]%>%
  # dplyr::slice(1:num_models)%>%
  # kbl(caption = "Table 1. Covariate Model Selection Results.",digits =3) %>%
  # kable_classic(full_width = F, html_font = "Cambria")


model_list<-lapply(best_covariates[[1]],#[best_covariates[[2]]$model_num[1:num_models]],
                   function(x) paste(x,collapse = " + "))%>%
  unlist()%>%
  as_tibble()%>%
  add_rownames()%>%
  dplyr::rename(model=rowname,model_name=value)


if(!file.exists("results/forecasts_2-3-26.rds")|redo_forecasts==TRUE){
  results<-SalmonForecasting::one_step_ahead(
    series=dat,
                  leave_yrs=leave_yrs,
                  TY_ensemble=TY_ensemble,
                  # covariates= best_covariates[[1]][best_covariates[[2]]$model_num[1:num_models]],
                  covariates=best_covariates[[1]],
                  first_forecast_period = first_forecast_period,
                  ts_freq = 1,
                  seasonal = FALSE,
                  plot_results = plot_results,
                  write_model_summaries = write_model_summaries,
                  forecast_period_start_m =  forecast_period_start_m, #inclusive 
                  forecast_period_start_d =  forecast_period_start_d, #inclusive
                  n_cores=8,
  include_mod = TRUE
                  )

save(results,file="results/forecasts_2-3-26.rds")
}else{
  load("results/forecasts_2-3-26.rds")
}

gc()


# contains_string <- function(lst, target) {
#   sapply(lst, function(vec) target %in% vec)
# }

limit_vars <- function(lst, num) {
  sapply(lst, function(vec) length(vec) <= num)
}
#
# target_string <- "lag1_PDO"
#
# includesPDO<-contains_string(best_covariates[[1]], target_string)
#
# results3_no_PDO <-results3[results3$model %in% as.character(which(!includesPDO)),]
# 
# lessthan<-limit_vars(best_covariates[[1]], 3)
# 
# results3_2 <-results3[results3$model %in% as.character(which(lessthan)),]
# nyr_check<-
# SalmonForecasting::check_nyrs(
#   One_ahead_out=results3_no_PDO,dat=dat,
# num_models=num_models,TY_ensemble=TY_ensemble,
#   model_list=model_list,
#   alpha=0,model_to_plot="MAPE_weighted"
# )

# nyr_check$plot_perf
# nyr_check$plot_pred

  
# alpha_check<-
# SalmonForecasting::check_alpha(
#   One_ahead_out=results,dat=dat,
# num_models=num_models,TY_ensemble=TY_ensemble,
# screen_metric="MAPE",
#   model_list=model_list,
#   roll_years=rolling_year_window,model_to_plot="MAPE_weighted"
# )

# alpha_check$plot_perf
#ggsave("alpha_perf.png")
# alpha_check$plot_pred
#ggsave("alpha_pred.png")
# 
# gridExtra::arrangeGrob(alpha_check$plot_perf+theme(axis.title.x = element_blank()), alpha_check$plot_pred, ncol = 1) |> ggsave(filename="alpha_both.png")

#calculate rolling performance
## rolling_year_window<-15
results<-results |> dplyr::select(-eq)

rp<-SalmonForecasting::rolling_perf(results,dat,rolling_year_window,screen_metric=MAPE,num_models,TY_ensemble,model_list,alpha=alpha)
# rp$performance

results<-results |> dplyr::select(-mod)


ens<-SalmonForecasting::ensemble(forecasts=rp$all_mods %<>% group_by(year) %>% mutate(rank=rank(MAPE)) %>% ungroup,
                                 series=dat,
                                 TY_ensemble=TY_ensemble,
                                 k=1,
                                 slide=rolling_year_window,
                                 num_models=num_models,
                                 stack_metric=stack_metric,
                                 stretch=FALSE,
                                 do_stacking = FALSE,
                                 alpha=alpha)
# ens$forecast_skill %>%filter(grepl("weight",model))

# weighted<-
# ens$ensembles |> filter(model=="MAPE_weighted") |> left_join(dat |> dplyr::select(year,abundance)) |> mutate(error=predicted_abundance-abundance) 

# rp$top_mods |> head(10) |> dplyr::select(predicted_abundance,model_name,MAPE,model,arma) |> left_join(ens$final_model_weights |> dplyr::select(model,MAPE_weight)) |>
#   dplyr::select(Prediction=predicted_abundance,MAPE,MAPE_weight,model=model_name,arma)|>
#     kbl(digits =2)%>%
#   kable_classic(full_width = F, html_font = "Cambria")
  
# rp$top_mods |> head(10) |> dplyr::select(predicted_abundance,eq,MAPE,model,arma) |> left_join(ens$final_model_weights |> dplyr::select(model,MAPE_weight)) |> 
#   dplyr::select(Prediction=predicted_abundance,MAPE,MAPE_weight,model=eq,arma)|> 
#     kbl(digits =2)%>%
#   kable_classic(full_width = F, html_font = "Cambria")


results_old<-evaluate_old_model(data=OPIHData,TY_ensemble=TY_ensemble,yr_end = yr_end)  

for_skill<-
rp$performance %>% mutate(model="Best_individual") %>% bind_rows(
ens$forecast_skill %>%filter(grepl("weight",model))
) %>% 
  bind_rows(results_old$forecast_skill %>% mutate(model="Current_OPI"))  %>%
  arrange(MAPE)



# top_mods<-one_step_ahead(series=dat |> mutate(across(all_of(covariates),\(x)c(scale(x)))),
#                   leave_yrs=1,
#                   TY_ensemble=1,
#                   # covariates= best_covariates[[1]][best_covariates[[2]]$model_num[1:num_models]],
#                   best_covariates[[1]][rp$top_mods |> filter(year==2026) |> pull(model) |> as.numeric()],
#                   first_forecast_period = first_forecast_period,
#                   plot_results = plot_results,
#                   write_model_summaries = write_model_summaries,
#                   forecast_period_start_m =  forecast_period_start_m, #inclusive
#                   forecast_period_start_d =  forecast_period_start_d, #inclusive
#                   stack_metric = stack_metric,
#                   k=1,
#                   include_mod = TRUE
#                   )
# 
# top_mod_coefs<-bind_rows(lapply(top_mods$mod,\(x)x[[1]])) |> mutate(model=rp$top_mods |> filter(year==2026) |> pull(model) ) |>
#   left_join(ens$final_model_weights |> dplyr::select(model,MAPE_weight)) |>
#   left_join(results |> filter(year==2026) |> dplyr::select(model,predicted_abundance))
#   
# 
# top_mod_coef_info<-
# top_mod_coefs |> mutate(across(dplyr::matches(covariates),\(x){x*MAPE_weight})) |> dplyr::select(dplyr::matches(covariates)) |>colSums(na.rm=TRUE)|> tibble::enframe("name","Average coefficient") |> arrange(desc(abs(`Average coefficient`))) |> left_join(
# 
# dat |> mutate(across(all_of(covariates),\(x)c(scale(x)))) |> filter(year==2026) |> dplyr::select(dplyr::matches(covariates))|>
# pivot_longer(everything(),values_to = "2026 covariate (Z-score)")
# ) |>
#   mutate("2026 covariate * ave. coeficient"=`Average coefficient`*`2026 covariate (Z-score)`)
# 
# 
# top_mod_coef_info |>
#   pivot_longer(`Average coefficient`:last_col(),names_to = "var",values_to = "value") |>
#   mutate(var=fct_relevel(var,c("2026 covariate (Z-score)","Average coefficient","2026 covariate * ave. coeficient"))) |>
# ggplot(aes(x=name,y=value))+geom_col()+facet_wrap(~var,scales="free")+theme(  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
#                                                                               axis.title = element_blank(),
#                plot.margin = margin(t = 0.5, r = .5, b = 0.5, l = 1, unit = "cm") # Add right margin
#                )
# 
# ggsave("cov_effects.png")


```








```{r results_stuff, message=FALSE, warning=FALSE, results="hide"}

Table2<-for_skill%>% mutate(model=sub("_"," ",model))%>% rename(Model=model)%>% 
  kbl(caption = paste0("Table 2. One step ahead individual and ensemble model performance"),digits =2)%>%
  kable_classic(full_width = F, html_font = "Cambria")




Table3<-
  # results$forecasts%>%
  # filter(year==max(year))%>%
  rp$top_mods %>% filter(rank==1) %>% arrange(year) %>% 
  dplyr::select(year,model_name,predicted_abundance,`Lo 50`,`Hi 50`,`Lo 95`,`Hi 95`) %>% #,Stacking_weight)%>%
     rename(Year=year,Model=model_name,`Predicted abundance` =predicted_abundance) %>% 
  kbl(caption = paste0("Table 3. Individual models which performed the best in",stack_metric," based on a",rolling_year_window, "year retrsopective analysis, and their forecasts."),digits =2)%>%
  kable_classic(full_width = F, html_font = "Cambria")


  
best_model<-"MAPE_weighted"#for_skill$model[1]


  results_best<-rp$top_mods %>% filter(model==best_model) %>% #mutate(model="MAPE_weighted") %>% 
  bind_rows(
  ens$ensembles %>% left_join(dat %>% dplyr::select(year,abundance)) %>% mutate(model_name=model)
) %>% filter(model==best_model) %>% 
  dplyr::select(year,model_name,abundance,
                  predicted_abundance,"Lo 50",
                "Hi 50","Lo 95","Hi 95") %>% 
  mutate(error=predicted_abundance-abundance,
         pct_errpr=error/abundance)






Table4<-results_best%>%
   rename(Year=year,Model=model_name,Abundance=abundance,`Predicted abundance`=predicted_abundance,`% error`=pct_errpr) %>% 
  # dplyr::select(!c("train_test","Stacking_weight","model_name") & !c(covariates[covariates%in%names(results_best)]))%>%
  kbl(caption = paste0("Table 4. One-year-ahead forecasts of OPI ", species," based on the ",best_model," model."),digits =2)%>%
  kable_classic(full_width = F, html_font = "Cambria")


Table4.1<-results_best%>%
   rename(Year=year,Model=model_name,Abundance=abundance,`Predicted abundance`=predicted_abundance,`% error`=pct_errpr) %>% 
  filter(Year==2026) |> dplyr::select(`Predicted abundance`:`Hi 95`) |> 
  # dplyr::select(!c("train_test","Stacking_weight","model_name") & !c(covariates[covariates%in%names(results_best)]))%>%
  kbl(caption = paste0("Forecast of 2026 OPI ", species," based on the ",best_model," model."),digits =2)%>%
  kable_classic(full_width = F, html_font = "Cambria")




Figure1<-ggplot(results_best,aes(x=year,y=predicted_abundance))+
ggplot2::geom_ribbon(aes(ymin=`Lo 95`,ymax=`Hi 95`),color=NA,alpha=0.5,fill = "cadetblue")+
  ggplot2::geom_ribbon(aes(ymin=`Lo 50`,ymax=`Hi 50`),color=NA,alpha=0.5,fill = "cadetblue")+
  geom_line()+
  geom_point(aes(x=year,y=abundance))+
  ylim(0,NA)+
  scale_x_continuous(breaks=seq(min(results_best$year),max(results_best$year),by=3))+ylab("Ocean abundance")+xlab("")

# ggsave(filename="fig1.png")

Figure2<-ggplot(rp$top_mods %>%
                  filter(rank==1) %>%
                  mutate(model="Best_individual") %>%
                  filter(year > (yr_end-TY_ensemble))%>%
                  bind_rows(
                    ens$ensembles %>%
                      left_join(dat %>%
                                  dplyr::select(year,abundance)) %>%
                      mutate(model_name=model))
                ,aes(x=year,y=predicted_abundance,col=model))+
  geom_line()+
  geom_line(data=results_best,mapping=aes(x=year,y=predicted_abundance),size=1.1,col="black")+
  geom_line(data=results_old$forecasts,mapping=aes(x=year,y=predicted_abundance),size=1.25,col="grey")+
  geom_point(aes(x=year,y=abundance),color="black")+
  ylim(0,NA)+
  scale_x_continuous(breaks=seq(min(results_best$year),max(results_best$year),by=3))+
  theme(legend.position = "none")+
  ylab("Ocean abundance")+xlab("")
  #theme(legend.key.size=unit(.15,'cm'),legend.position = "top")

# ggsave(filename="fig2.png")

Figure3<-ggplot((results_old$forecasts%>%mutate(model_name="Current_OPI") %>% bind_rows(results_best ) %>% mutate(Model=sub("_"," ",model_name)) %>%  mutate(pct_error=log10(((predicted_abundance-abundance)/abundance)+1))) %>% filter(year<yr_end),
                  aes(x=year,y=pct_error,color=Model))+geom_hline(yintercept=0)+geom_line(lwd=1.25)+scale_color_manual(values=c("darkblue","darkred"))+xlab("")+ylab(expression(paste(log[10], "(% error +1)")))

# ggsave(filename="fig3.png")
```





## Results

<!-- Table 1 shows the model selection results sorted by AIC based on in-sample fits of ARIMA models tested on all subsets of covariates to the full (all years) data set. The ARIMA model structure for each covariate set was selected by the auto.arima function in the forecast package. These model selection results are generated as a first step to identify the top models for further use in evaluating out-of-sample one-step-ahead forecast skill, and constructing ensembles. -->

<!-- ```{r message=FALSE, warning=FALSE, results="show"} -->

<!-- Table1 -->

<!-- ``` -->

Table 2 shows the model selection results sorted by `r stack_metric`
based on `r TY_ensemble-1` years of out-of-sample one-step-ahead
leave-one-out cross validation. Results include both individual ARIMA
models as well as weighted ensembles. Ensembles include both a "stacking
weight" ensemble where an optimization algorithm is used to identify the
weights that maximize out-of-sample one-step-ahead performance over a
sliding `r 15` year window using the selected performance metric for the
training years, as more traditional ensembles using a harmonic mean
where weight $w$ may be calculated for model $i$ from model set $K$
based on it's skill $s$ versus that of other models in the set, where
$s$ may be measured by RMSE, MASE, MAPE, MSA, etc:
$${w}_{i} = \frac {{s}_{i}^{-1}} {\sum_{i=1}^{K} {s}_{i}^{-1}} $$

```{r message=FALSE, warning=FALSE, results="show"}
Table2
```

Table 3 shows predictions of total run size for the individual model that
would have been selected in each year, as well as which model would have
been selected.

```{r message=FALSE, warning=FALSE, results="show"}
Table3
```

Table 4 and Figure 1 shows annual one-step-ahead predictions of
abundance for the best individual model, including 50 and 95% prediction intervals
relative to the observed abundance (black dots in Figure 1).

```{r message=FALSE, warning=FALSE, results="show"}
Table4
```

```{r message=FALSE, warning=FALSE, results="show", fig.align='center',fig.show=TRUE, fig.cap="Figure 1. One-step-ahead predictions (black line) of total OPI coho abundance for the best model relative to observed abundance (black dots). Prediction intervals are shown as blue shading including 50% (dark), 95% (light)"}
Figure1
```

Figure 2 shows the annual one-step-ahead predictions of abundance
(lines) for the entire model set and the observed abundance (black
dots), the MAPE averaged forecast (thick black line), and the existing OPI
forecast (thick grey line).

```{r message=FALSE, warning=FALSE, results="show", fig.align='center',fig.show=TRUE, fig.cap="Figure 2. One-step-ahead predictions (lines) of total OPI coho abundance for the top model set relative to observed abundance (black dots)."}
Figure2
```

Figure 3 shows the log transformed percent forecast error of the current and new model over time.

```{r message=FALSE, warning=FALSE, results="show", fig.align='center',fig.show=TRUE, fig.cap="Figure 3. The log transformed percent forecast error of the current and new model over time."}
Figure3
```





Figure 4 shows time series of the z-scored covariate values.

```{r message=FALSE, warning=FALSE, results="show", fig.align='center',fig.show=TRUE, fig.cap="Figure 6. Standardized covariate values by year of forecast for most recent 10 years."}

integer_breaks <- function(n = 5, ...) {
  fxn <- function(x) {
    breaks <- floor(pretty(x, n, ...))
    names(breaks) <- attr(breaks, "labels")
    breaks
  }
  return(fxn)
}


dat %>% dplyr::select(all_of(covariates),year) %>% mutate(across(all_of(covariates),function(x)c(scale(x))),year=as.integer(year)) %>% pivot_longer(all_of(covariates),values_to = "Z-scored value") %>% filter(between(year,2000,2026)) %>% ggplot(aes(x=year,y=`Z-scored value`))+geom_point()+geom_line() + facet_wrap(~name)+geom_hline(yintercept = 0,lty=2)+scale_x_continuous(breaks=integer_breaks(),guide = guide_axis(check.overlap = TRUE))

```



## Appendix

Appendix 1 shows the raw data and covariates used in modeling.

```{r message=FALSE, warning=FALSE, results="show"}
dat%>%
  dplyr::select(-period,-`...1`)%>%
  kbl(caption ="Appendix 1. data used for forecasting OPI Coho",digits =2)%>%
  kableExtra::kable_classic(full_width = F, html_font = "Cambria")
```



OPI data and covariate notes: 1) Jack OPI = Total Jack CR and Jack OC.
2) Jack CR = Columbia River jack returns corrected for small adults. 3)
Jack OC = Oregon coastal and California hatchery jack returns corrected
for small adults. 4) Total OPI = Columbia River (Sm D + Sm CR), Oregon
coastal and Klamath Basin. 5) Sm CR = Columbia River smolt releases from
the previous year expected to return as adults in the year listed. 6) Sm
D = Columbia River delayed smolt releases from the previous year
expected to return as adults in the year listed 7) SmAdj = JackCR\*(Sm
D/Sm CR) (this is a covariate used in the past OPI forecast)

OCN covariates 1) Multivariate ENSO Index (Oct-Dec; t-1) 2) Spring
Transition (Julian date; t-1) 3) Upwelling (July-Sept; t-1) 4) Upwelling
(Sep-Nov; t-1) 5) Sea Surface Temperature (Jan; t) 6) Sea Surface Height
(April-June; t-1) 7Z)PDO (mean May-July; averaged over the 4-years prior
to the return year)

Ancillary Covariates 1) [Pacific Decadal
Oscillation](https://www.ncei.noaa.gov/access/monitoring/pdo/) 2) [North
Pacific Gyre Oscillation](http://www.o3d.org/npgo/) 3) [NOAA Ocean
Indicators](https://www.fisheries.noaa.gov/west-coast/science-data/ocean-ecosystem-indicators-pacific-salmon-marine-survival-northern)
4) [NOAA NOAA Extended Reconstructed SST version
5](https://psl.noaa.gov/data/gridded/data.noaa.ersst.v5.html) 5) [NOAA
Oceanic Nino 3.4 Index (lag 1 Oct-December
mean)](https://psl.noaa.gov/gcos_wgsp/Timeseries/Nino34/)
